---
title: "One-way ANOVA and chi-square examples"
author: "Allison Horst"
date: "11/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(car) # Used for leveneTest for equal variances in ANOVA
```

## One-way ANOVA

#### Make some mock data 
```{r}
set.seed(1020)
df_aov <- data.frame(treatment = rep(c("A","B","C"), each = 20),
                     value = c(rnorm(mean = 5, sd = 1.3, n = 20), 
                               rnorm(mean = 8, sd = 1.8, n = 20),
                               rnorm(mean = 6, sd = 1.5, n = 20)))
```

#### Explore data & check assumptions

Visual exploration: 
```{r}
ggplot(data = df_aov, aes(x = value)) +
  geom_histogram(bins = 12) +
  facet_wrap(~treatment, scales = "free")

ggplot(data = df_aov, aes(sample = value)) +
  geom_qq() +
  facet_wrap(~treatment, scales = "free")

# Groups appear ~ normally distributed
# If large n, means to be normally distributed regardless of underlying population (CLT)
```

Check variances: 
```{r}
# Create a summary table (including variances) by group
df_summary <- df_aov %>% 
  group_by(treatment) %>% 
  summarize(
    gp_mean = mean(value),
    gp_se = sd(value) / sqrt(n()),
    gp_var = var(value)
  )

df_summary # Largest variance (5.62), smallest variance (1.76)

# Levene's statistical test for equal variance: 
leveneTest(value ~ treatment, data = df_aov) # There IS a significant difference
# ...but, largest variance < 4x smallest variance
```

#### Run one-way ANOVA to compare group means

```{r}
my_aov <- aov(value ~ treatment, data = df_aov)
summary(my_aov) 
# There IS a significant difference in means across groups. Which ones? 

# Run post-hoc Tukey's:
my_post_hoc <- TukeyHSD(my_aov)
my_post_hoc
# All pairwise combinations here are significantly different

```

#### Reporting outcomes of one-way ANOVA with post-hoc Tukey's in text: 

```{r}
# Getting the outputs from that ANOVA summary are a bit different.
# There are many ways to pull the information from a summary of your ANOVA.
# I find this way easiest. 

# Outcomes of summary(my_aov) are stored in a list. Save the unlisted version: 
aov_outputs <- unlist(summary(my_aov))

# Look at aov_outputs:
aov_outputs

# The order that things appear in aov_outputs is the index number you'll reference to pull those pieces. 
# Here are the pieces you'll need:

# The first item [1] is the numerator degrees of freedom; 
# The second item [2] is the denominator degrees of freedom; 
# The seventh item [7] is the F-statistic value;
# The ninth item [9] is the p-value

```
Pull those items as follows to create a statement of the ANOVA results (see Rmd for in-line referencing): 

Mean phosphate concentration (mg/L) differed significantly across all three groups (one-way ANOVA with post-hoc Tukey's HSD, F(`r aov_outputs[1]`, `r aov_outputs[2]`) = `r round(aov_outputs[7],2)`, *p* = `r round(aov_outputs[9],3)`).

**Note:** In that case, you could just have p < 0.001 (since rounding will say p = 0). 

**Note:** That is ONLY the statement of statistical outcomes. It should be accompanied by a more meaningful comparison of differences, e.g. by showing the groups visually or in a table with a measure of data spread/uncertainty, and by describing actual differences in text. 

## Chi-square test for independence

#### Make a mock contingency table: 
```{r}

# The counts of cats (top row) and dogs (bottom row) who choose pizza or burritos as their favorite food: 

df_chi <- tribble(
  ~pizza, ~burrito,
  54, 30,
  40, 48
)

df_chi
```


#### Chi-square test for independence

**We ask:** Does favorite food (pizza/burrito) differ significantly by species? 
Chi-square! 

- See .Rmd for how to create the actual symbol here: ($\chi^2$)

Run the test: 
```{r}
my_chi_test <- chisq.test(df_chi)
my_chi_test

# The chisq.test outputs are similar to the t.test outputs
# See the 'Value' section for ?chisq.test
```

#### Reporting the outcomes of chi-square in text (see Rmd for how to write this with in-line referencing): 

There is a significant effect of species (cat/dog) on food preference (pizza/burrito) ($\chi$^2^(`r my_chi_test$parameter`) = `r round(my_chi_test$statistic,2)`, *p* = `r round(my_chi_test$p.value, 2)`). 

**Note:** This is only a statement of statistical results, and should be a piece of a more interesting and meaningful discussion about the actual proportions of cats and dogs choosing pizza vs. burritos. 



