---
title: 'ESM 206 Lab 5'
author: "Allison Horst"
date: "10/25/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Counts, uncounts, t-tests, ungroup, lubridate

#### Lab 5 objectives: 

- Get counts of grouped observations using `count()`
- `ungroup()` grouped data (and know why that can be essential) 
- Perform one and two sample t-tests with `t.test()`
- Parse dates with `lubridate`
- Build a `geom_tile()` heatmap

Set-up: Students should Fork & Clone the **allisonhorst/esm-206-f2019-lab-5** repo from GitHub prior to the lab to work with materials locally. 

### Lab 5 data: SBC LTER Lobster abundance and fishing pressure

**Link:** https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=77&revision=newest

**Citation:** Reed D. 2019. SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012. Environmental Data Initiative. https://doi.org/10.6073/pasta/a593a675d644fdefb736750b291579a0. Dataset accessed 10/25/2019.

#### 1. Attach packages

Required: `tidyverse`, `here`, `janitor`
```{r, message = FALSE, warnings = FALSE}
library(tidyverse)
library(here)
library(janitor)
```

#### 2. Read in data

- **lobster_abundance.csv**
- **lobster_traps.csv**

```{r, message = FALSE}
lobster_abundance <- read_csv(here("lab_5_materials",
                                   "data",
                                   "lobster_abundance.csv"),
                              na = "-99999") %>% 
                              clean_names()
```

First, let's ask: How is *lobster_abundance* not in tidy format? 

- Each observation (lobster) is not in it's own row. (e.g. sort by `lobster_count` from high to low: there are sometimes many lobsters represented in a single row!). 

- This is called a **frequency table**, but we want this to be in **case format** - so we need to expand the data frame so that each lobster is in its own row in order to be able to do analyses. 

We can use `tidyr::uncount()` to expand a df by a frequency variable (here, by the value in `lobster_count`).

#### 3. Use `tidyr::uncount()` to convert to tidy format (one observation per row)

```{r}
lobster_tidy <- lobster_abundance %>% 
  tidyr::uncount(lobster_count)
```

Now each lobster observed has been placed into its own *row* - which is good, because now we don't have to worry about weights when we find summary statistics. 

#### 4. Always look at your data. Always. 

Various by site (the variable we'll consider later on):  
```{r}

# Jitterplot
ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site)) 

# Violin plot: 
ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_violin()

# Histogram:
ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram() +
  facet_wrap(~site)

# Quantile-quantile plot: 
ggplot(lobster_tidy, aes(sample = size_mm)) +
  geom_qq() +
  facet_wrap(~site)

```


#### 5. Parse dates for lobster observations

Notice that the existing class of the `date` variable is a character. Boo.
```{r}
class(lobster_abundance$date)
```

The `lubridate` package (part of the `tidyverse`), and is built to make it easier to deal with date-time data. Here, we'll use `lubridate::mdy()` to help R understand it's a date, and work with the pieces (month, day and year) more easily. 

**Note**: we're using `mdy()` because that's the existing format of the date column (mm/dd/yy). Check `?ymd` to see all the different options based on the format of your date column. Cool. 

Add a new column with `mutate()` that is a *date* using `mdy()`: 
```{r}
lobster_date <- lobster_tidy %>% 
  mutate(
    date_new = lubridate::mdy(date)
  )
```

Check it out! The *date_new* column is in nice ISO date format. Let's check the class:
```{r}
class(lobster_date$date_new)
```

Wooooo.

Now that it's in date format, we can parse it. See `?month` and `?year` - to get or set components of a date or date-time. So here, we'll use `lubridate::month()` and `lubridate::year()` to create separate columns for the month and year. 

For `month()`, we can even automatically convert to month abbreviation with argument `label = TRUE`. 

```{r}
lobster_parse_date <- lobster_date %>% 
  mutate(obs_month = lubridate::month(date_new, label = TRUE),
         obs_year = lubridate::year(date_new))
```

A really cool thing about that? When we use `lubridate::month()`, it's auto stored as an ordered factor booya (so we don't need to `fct_relevel()`). 

Now let's find some counts of lobster observations using different groupings. 

#### 6. Find counts by groups with `dplyr::count()`

There are a bunch of ways to count things based on different variables in R. So many that sometimes it's challenging to know which to use. We'll touch on a few of the most generally useful ones here. 

- `dplyr::count()`: groups, tallies, then ungroups (whoa)
- `dplyr::tally()`: use `group_by()` first, then use `tally()` for counts, then manually need to `ungroup()`
- `dplyr::n()`: number of observations in current group

##### `count()` example 1: Count the number of lobsters by year and month:

```{r}
lobster_ym <- lobster_parse_date %>% 
  count(obs_year, obs_month)

lobster_ym
```

##### `count()` example 2: Count the number of lobsters by year and site:

```{r}
lobster_ysite <- lobster_parse_date %>% 
  count(obs_year, site)

lobster_ysite
```

##### `count()` example 3: Count the number of lobsters by site: 

```{r}
lobster_site <- lobster_parse_date %>% 
  count(site)

lobster_site
```

##### For comparison: group first, then use `tally()` to do the same thing. Other alternatives: `add_tally()`, `add_count()` for mutate versions. 
```{r}
lobster_tally <- lobster_parse_date %>% 
  group_by(site) %>% 
  tally()

lobster_tally
```

What if we want to make a summary table that has a mean, sd, and the sample size? Use `dplyr::n()` to find the number of observations in the current group. 

```{r}
lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    sample_n = n()
  )

lobster_summary
```

So, to reiterate - three ways to do the same thing to get sample sizes: 
```{r, eval = FALSE}
lobster_n <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    n = n()
  ) %>% 
  ungroup()

lobster_tally <- lobster_parse_date %>% 
  group_by(site) %>% 
  tally() %>% 
  ungroup()

lobster_count <- lobster_parse_date %>% 
  count(site)

```

Hooray! 

More information on tally/count: see descriptions and examples from [tidyverse.org](https://dplyr.tidyverse.org/reference/tally.html). 

#### 7. One- and two-sample t-tests with `t.test()`

So we found the mean and sd of lobster sizes (see object *lobster_summary*). But what if we ask the question: Is there a significant difference in lobster sizes between Naples Reef and Carpinteria Reef? 

We've already looked at the data (jitterplots, histograms, etc.) and it seems like a t-test would be an appropriate test. Ask here: why? Sample size, distribution, data type, etc....

OK, so we've decided that a t-test is appropriate and useful. How do we run a t-test in R? 

##### Method 1: Isolate vectors of observations, then use `t.test()`

First, we could get the vectors of observations from the two groups we want to compare, then add them as arguments to `t.test()`. Like this: 

```{r}
# Get a vector only containing size observations (mm) for lobsters at Naples Reef (site == "NAPL"):

napl_sample <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

# Get a vector only containing size observations (mm) for lobsters at Mohawk Reef (site == "MOHK"): 

mohk_sample <- lobster_tidy %>% 
  filter(site == "MOHK") %>% 
  pull(size_mm)
```

Check out those two objects. Note that they are *vectors* only containing size observations for lobsters at the two sites. 

We can add those as our first two arguments in `t.test()` to run a two-sample t-test:
```{r}
mohk_napl_ttest <- t.test(mohk_sample, napl_sample)
mohk_napl_ttest
```

Interpret the outcome. Now let's say we want to write a statement about the result (for now, in very "stats class" language, which we'll ditch quickly):

Mean lobster size differed significantly between Mohawk and Naples Reefs (t(`r round(mohk_napl_ttest$parameter, 2)`) = `r round(mohk_napl_ttest$statistic, 2)`, *p* < 0.001). 

**Remember**: the more important things are the data itself, means comparison, effect size, etc. - which we'll also get to. 

##### Method 2: Filter to only include 2 groups, then use `t.test(A ~ B)`

```{r}
# First, filter to only include 2 groups (MOHK & NAPL):

lobster_2sample <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

# Then:

ttest_2 <- t.test(size_mm ~ site, data = lobster_2sample)
```

And similarly use in-line referencing to add text directly from coded objects. Woo reproducibility! 

#### END LAB.
