---
title: "ESM 206 Lab 3"
subtitle: "Reading in Excel files, exploratory & final dataviz, intro to `here::here()`, exporting figures, `dplyr::case_when()`, `purrr::modify_if()`"
author: "Allison Horst"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### Lab 3 Objectives:

- R Markdown continued
- Meet the `here` package for better file paths
- Reading in data from a text file 
- Write data to a CSV file
- Exploratory & final dataviz (jitter, scatter, violin, ridgeline, polar coordinates)
- Saving figures with `ggplot2::ggsave()`
- Intro to some loops with `purrr`
- If-else statements with `dplyr::case_when()`


### Data: NOAA Santa Barbara buoy (Station ID NTBC1) data, 2018

**Source:** https://www.ndbc.noaa.gov/station_history.php?station=ntbc1
**Data from:** National Buoy Data Center (NOAA), accessed 10/4/2019
**Data descriptions:** https://www.ndbc.noaa.gov/measdes.shtml

## Part 0. Set-up

- Create a new R project 'esm_206_lab_3' within your labs folder
- Within the project directory, add a new subfolder *raw_data*
- Create a second subfolder called *intermediate_data*
- Drop the text file 'sb_buoy_2018.txt' into the *raw_data* subfolder

## Part 1. Attach packages


**NOTE:** need to have {hexbin} installed previously, even though not attached here, for some to use geom_hex()
```{r}
library(tidyverse)
library(here)
library(janitor)
library(ggridges)
```

## Part 2. Read in data from a text file (sb_buoy_2018.txt)

Use `here::here()` to let R know where to find the data!
And use `readr::read_table()` to "read whitespace-separated columns into a tibble." 

```{r}

sb_buoy <- readr::read_table(here::here("lab_3_materials",
                                 "raw_data", 
                                 "sb_buoy_2018.txt"),
                             na = c("99", "999", "99.0", "999.0")) %>% # Note: students will just have "raw_data" here
  janitor::clean_names() %>% 
  dplyr::slice(-1) %>% # Use dplyr::slice() to remove or keep rows by position
  select(number_yy:gst, atmp)

```

## Part 3. Write a file to a subdirectory with `here::here()`

Now if I want to write an intermediate file (to keep this cleaned up version...)

```{r}
write_csv(sb_buoy, here("lab_3_materials", "intermediate_data", "sb_buoy.csv"))
```

## Part 4. Make the 'month' column ordered factor with month names

First, check it out:
```{r}

class(sb_buoy$mm) # It's a character
unique(sb_buoy$mm) # All the values here

# Check out month.abb() built-in vector
month.abb # Cool - pre-stored month abbreviates

sb_buoy_month <- sb_buoy %>% 
  mutate(mm = as.numeric(mm)) %>% # Convert month to numeric %>% 
  mutate(
    month_name = month.abb[mm] # Note square brackets here
  ) 

# Check the class of month_name
class(sb_buoy_month$month_name)

# But that means if we plot it, R will just do it alphabetically. 
# We want to be able to assign these an *order*
# So we need to make it a factor, and specify the factor levels (Jan, Feb, etc.)
# Use mutate() to overwrite the column (carefully...)

sb_buoy_fct <- sb_buoy_month %>% 
  mutate(
    month_name = fct_relevel(month_name, levels = month.abb)
  ) 
### could also arrange(as.numeric(mm)) %>% mutate(month_name = fct_inorder(month_name))
### -- this method would work for (e.g.) the country salmon production 
### data where the order is not known in advance.

levels(sb_buoy_fct$month_name)
# Now R will understand that order matters, & will plot accordingly

```

## Part 5. `purrr` functions for loops, and some dataviz

First, let's explore air temperatures (atmp) by month:
```{r}

ggplot(sb_buoy_fct, aes(x = month_name, y = atmp)) +
  geom_jitter() # Weird. Why? 

# Always check classes of things: here, notice that all are read in as character. 
# Not great. Let's convert multiple columns to "numeric" at once using purrr::map()
# purrr::map

```

YUCK that's wrong. 

Check the class of the variables with `summary()`. 

We see that these "values" are stored as characters, when the need to be values. One option is to use as.numeric() for each column separately. Or we can use `purrr::modify_if()` to apply a function to *multiple* variables that meet a condition by looping over them. 

```{r}

sb_buoy_num <- sb_buoy_fct %>% 
  purrr::modify_if(is.character, as.numeric)

```

Now let's try looking at the air temperatures again: 
```{r}

ggplot(sb_buoy_num, aes(x = month_name, y = atmp)) +
  geom_jitter() # Too many points, but trends clear

ggplot(sb_buoy_num, aes(x = month_name, y = atmp)) +
  geom_violin() # A little hard to compare still, but good

ggplot(sb_buoy_num, aes(x = atmp)) +
  geom_density(aes(fill = month_name),
               color = NA,
               show.legend = FALSE) + # Cool but useless
  facet_wrap(~month_name) +
  theme_light()

```

Hmmm but that's hard to see because of order, even if we facet_wrap(). Here are a couple of better options: 

### Option 1: overlay all single month plots on top of the *overall* plot for data from all months so that there is a "reference population" for each month: 

- How? We need to add a layer where the faceting variable is set to `NULL`.
- Here, we'll make a histogram: 

```{r}

ggplot(sb_buoy_num, aes(x = atmp)) +
  geom_histogram(data = transform(sb_buoy_num, month_name = NULL), fill = "gray90") +
  geom_histogram(aes(fill = month_name),
               color = NA,
               show.legend = FALSE) + # Cool but useless
  facet_wrap(~month_name) +
  theme_light()

ggsave(here("lab_3_materials", "figures", "temp_hist.png"), height = 6, width = 6)


```

### Option 2: a ridgeline plot with `ggridges`

```{r}

temp_graph <- ggplot(sb_buoy_num, aes(x = atmp, y = month_name)) +
  geom_density_ridges(fill = "gray60",
                      color = "gray10",
                      size = 0.2) +
  scale_x_continuous(lim = c(5,25)) +
  theme_minimal() +
  labs(x = "Air temperature (Celsius)",
       y = "Month (2018)",
       title = "SB buoy monthly temperatures (2018)",
       subtitle = "Source: NOAA Nationa Buoy Data Center") +
  scale_y_discrete(limits = rev(levels(sb_buoy_num$month_name))) # rev months

temp_graph

```

What if I wanted to save this? 

```{r}

ggsave(here("lab_3_materials", "figures", "temp_graph.png"), height = 6, width = 6)
# Note: you can update size, resolution, etc. within ggsave
```

Let's explore a relationship: 

Windspeed vs. wind direction?
```{r}

# A few different graph types: 

# Plain scatterplot
ggplot(sb_buoy_num, aes(x = wdir, y = wspd)) +
  geom_point(aes(color = wspd)) 

# 2d density plot
ggplot(sb_buoy_num, aes(x = wdir, y = wspd)) +
  geom_density_2d()

# Hex density plot
ggplot(sb_buoy_num, aes(x = wdir, y = wspd)) +
  geom_hex(bins = 50) +
  scale_fill_gradient(low = "orange", high = "red")

# Or, use scale_fill_gradientn(colors = c("white","yellow","orange","purple"))

# Finalize hex density and wrap by month
ggplot(sb_buoy_num, aes(x = wdir, y = wspd)) +
  geom_hex(bins = 30) +
  scale_fill_gradientn(colors = c("yellow","orange","purple")) +
  scale_y_continuous(lim = c(0,10), expand = c(0,0)) +
  scale_x_continuous(lim = c(0,360), expand = c(0,0)) +
  theme_dark() +
  facet_wrap(~month_name) # See how it changes by month? 

```

It does look like there is some pattern to the windspeeds (like, we rarely have strong winds coming from ~180 degrees, but a lot of strong winds coming from ~240 degrees). Might make sense here to plot on a polar coordinate system: 

```{r}

ggplot(sb_buoy_num, aes(x = wdir, y = wspd)) +
  geom_density_2d(aes(color = month_name),
                  size = 0.2,
                  show.legend = FALSE) +
  coord_polar() +
  scale_x_continuous(breaks = c(0, 90, 180, 270), labels = c("N","E","S","W")) +
  facet_wrap(~month_name) +
  theme_minimal() +
  labs(x = "", y = "windspeed (mph)", title = "SB windspeed and direction (2018)")

ggsave(here("lab_3_materials", "figures", "wdir_polar.jpg"))

```

## Part 6. `dplyr::case_when()` for easier if-else statements

Let's say that we actually want to just explore things on a *seasonal* level, where:

- Spring = Mar, Apr, May
- Summer = Jun, Jul, Aug
- Autumn = Sep, Oct, Nov
- Winter = Dec, Jan, Feb

Here, we'll make a new column that contains the *season* associated with each observation using `dplyr::mutate()` + `dplyr::case_when()` 
```{r}

sb_buoy_season <- sb_buoy_num %>% 
  dplyr::mutate(
    season = dplyr::case_when(
      month_name %in% c("Mar", "Apr", "May") ~ "spring",
      month_name %in% c("Jun", "Jul", "Aug") ~ "summer",
      month_name %in% c("Sep", "Oct", "Nov") ~ "autumn",
      month_name %in% c("Dec", "Jan", "Feb") ~ "winter"
    )
  )

```

Then we could make calculations by season, for example: 

What is the mean windspeed by season for 2018? 
```{r}

mean_wind <- sb_buoy_season %>% 
  group_by(season) %>% 
  summarize(
    mean_wspd = mean(wspd)
  )

mean_wind

```

### END LAB. 
