---
title: "ESM 206 Fall 2019 - Lab 10"
author: "Allison Horst"
date: "11/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(corrplot)
library(beepr)
library(praise)
library(stargazer) # For nice regression tables! 
# Hlavac, Marek (2018). stargazer: Well-Formatted Regression and Summary Statistics Tables.
#  R package version 5.2.2. https://CRAN.R-project.org/package=stargazer 
library(sf)
library(gganimate)
library(transformr)
```

## Objectives:

- Multiple linear regression
- Check assumptions w/diagnostic plots
- Make predictions with new data
- Some other fun things (beepr, praise, sf, gganimate)

### 1. Multiple linear regression

Here, we'll be modeling housing price based on house characteristics using data from SLO home sales.

- Read in slo_homes.csv file and explore
- Only keep observations for San Luis Obispo, Atascadero and Arroyo Grande

```{r}
homes <- read_csv("slo_homes.csv") %>% 
  clean_names() %>% 
  filter(city %in% c("San Luis Obispo", "Atascadero", "Arroyo Grande"))
  
praise()
```

Some exploring: look at correlations between numeric variables
```{r}
homes_cor <- cor(homes[2:6])
homes_cor

corrplot(homes_cor, 
         method = "circle",
         type = "upper")

# Moderate correlations between numeric variables (no concern about multicollinearity here from a correlation value standpoint, but there may be some conceptual overlap between variables)

praise()
# See: names(praise_parts) for other things you can call in praise

# For example, customize it a bit
praise("You are totally ${adjective}! Super ${EXCLAMATION}!")
```

But let's start out using all variables we have: 
```{r}

home_lm <- lm(price ~ city + bedrooms + bathrooms + sq_ft + status, data = homes)

# View it: 
home_lm
summary(home_lm)

```

Well that's kind of a nightmare to look at. And putting it into a table could be really challenging. Enter, the `stargazer` package. 

```{r, results = "asis"}
stargazer(home_lm, type = "html")
```

Let's answer a few questions: 

- How do we interpret each of these variables? 
- What is the reference level for city? 
- What does NOT make sense here, that might be multiply described by another variable in the model?

Try another version of the model:
```{r}
home_lm2 <- lm(price ~ city + sq_ft + status, data = homes)
```

And check out the results: 
```{r, results = "asis"}
stargazer(home_lm2, type = "html") # And you can customize...
```

You can also use stargazer for multiple model comparisons: 
```{r, results = "asis"}
stargazer(home_lm, home_lm2, type = "html")
```

### 2. Exploring diagnostics

We can use the diagnostic plots to check assumptions about residuals, e.g.:

- Constant variance of residuals?
- Normally distributed residuals?
- Also check: any notable outliers? 

```{r}
plot(home_lm)
```

### 3. Predictions for home price with multiple variables 

First, we'll make a new data frame containing all variables that `home_lm2` needs to make a prediction: 

```{r}
new_df <- data.frame(
  city = rep(c("San Luis Obispo", "Arroyo Grande", "Atascadero"), each = 10),
  sq_ft = rep(seq(0, 5000, length = 10)),
  status = rep("Regular", 30)
)

new_df
```

Then, use the `predict()` function to find the predicted home prices for each combination in `new_df`:

```{r}

# Make the predictions using new_df:
predict_df <- predict(home_lm2, newdata = new_df)

# Then bind predictions together with new_df:
full_df <- data.frame(new_df, predict_df)

```

### 4. Visualize it! 

```{r}
ggplot() +
  geom_point(data = homes, 
             aes(x = sq_ft, 
                 y = price,
                 color = city,
                 pch = city),
             size = 1,
             alpha = 0.5) + 
  geom_line(data = full_df,
            aes(x = sq_ft, 
                y = predict_df,
                color = city)) +
  scale_color_manual(values = c("orange", "magenta", "black")) +
  theme_light()

```

### 5. Compare AIC values 

...but statistics are not substitute for judgement!
```{r}
AIC(home_lm)
AIC(home_lm2)

# Pretty close, but the first model has a lower AIC. Should I pick the first model just based on AIC? NO!
```

### 6. A little map teaser

Data source: 
[https://nid.sec.usace.army.mil/ords/f?p=105:19:14606625072661::NO:::](National Inventory of Dams)

Get and check out the CA dams data: 
```{r}
dams <- read_csv("ca_dams.csv") %>% 
  clean_names() %>% 
  drop_na(latitude) %>% 
  drop_na(longitude) %>% 
  drop_na(year_completed)
```

Then make sure R understands that latitude & longitude are spatial coordinates using `sf::st_as_sf()`:
```{r}
dams_sf <- st_as_sf(dams, 
                    coords = c("longitude","latitude")
                    )

st_crs(dams_sf) <- 4326

# Check class:
class(dams_sf)
```

What does that mean? Now R recognizes this as spatial data. 

```{r}
plot(dams_sf)
```

Cool! Now let's get an outline for California: 

# NOTE: GET SAME PROJECTIONS HERE! 
```{r}
ca_border <- read_sf(here::here("lab_10_materials", "ca_state_border"), layer = "CA_State_TIGER2016")

plot(ca_border)
```

Then plot them together with ggplot2!
```{r}

ggplot() +
  geom_sf(data = dams_sf)

ggplot() +
  geom_sf(data = ca_border)

# Combine: 
ggplot() +
  geom_sf(data = ca_border) +
  geom_sf(data = dams_sf,
          size = 1,
          alpha = 0.5,
          color = "black") +
  theme_minimal()

beepr::beep(8)
```

### 7. A little gganimate teaser

Now show how dams have been added over time with `gganimate`!
```{r}
ggplot(data = dams_sf) +
  geom_sf(data = ca_border) +
  geom_sf(data = dams_sf, 
          size = 1.5,
          color = "gray50") +
  theme_void() +
  labs(title = 'Year: {round(frame_time,0)}') +
  transition_time(year_completed) +
  shadow_mark(alpha = 1)
