---
title: "Beaver body temperatures t-test"
author: "Allison Horst"
date: "11/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
```

First, explore distributions (even if you're not asked to):
```{r}
ggplot(data = beaver1, aes(x = temp)) +
  geom_histogram()

ggplot(data = beaver1, aes(sample = temp)) +
  geom_qq()

ggplot(data = beaver2, aes(x = temp)) +
  geom_histogram()

ggplot(data = beaver2, aes(sample = temp)) +
  geom_qq()

# Beaver 2 may not be normally distributed, but sample sizes are large (both > 100), so Central Limit Theorem: means will be normally distributed, & we can proceed with a parametric test.

# Also look at them side by side: 

ggplot() +
  geom_histogram(data = beaver1, aes(x = temp),
                 fill = "red") +
  geom_histogram(data = beaver2, aes(x = temp),
                 fill = "blue")
```

Perform a t-test (a couple of different ways):

1. Is there a significant difference in body temperatures between beaver 1 and beaver 2?
```{r}

beaver_diff <- t.test(beaver1$temp, beaver2$temp)

```

In-line report: See ?t.test 'values' section

Mean body temperatures (^$\circ$^C) for beaver 1 (`r round(beaver_diff$estimate[1],2)`) and beaver 2 (`r round(beaver_diff$estimate[2],2)`) differ significantly (t(`r round(beaver_diff$parameter,2)`) = `r round(beaver_diff$statistic,2)`, *p* = `r round(beaver_diff$p.value,2)`). 

**Consider just updating to p < 0.001 for very small p-values.**

Another question: Is beaver 2 mean body temperature significantly higher than beaver 1? 

```{r}
beaver2_greater <- t.test(beaver2$temp, beaver1$temp, alternative = "greater")
```

Mean body temperature for beaver 2 is significantly greater than that for beaver 1 (t(`r round(beaver2_greater$parameter,2)`) = `r round(beaver2_greater$statistic,2)`, *p* = `r round(beaver2_greater$p.value,2)`). 
